CLEAN DATASET

Load packages:

```{r, message=FALSE}
library(stringr)
library(ggplot2)
library(raster)
library(dplyr)
library(downloader) # for source_url()
library(rgdal) #for readOGR()
library(gdm) # for GDM analyses
library(WhereNext)
library(DHARMa)
library(glmmTMB)
```

Load data:

```{r}
url <- "https://github.com/ibartomeus/IberianBees/raw/master/Data/iberian_bees.csv.gz"

download.file(url = url,
              destfile = "./Data/iberian_bees.csv.gz") # download date = 17/7/2022

data <- read.csv(gzfile("./Data/iberian_bees.csv.gz"))

# rename some columns so the dataset can be properly read by the clean_coordinates function in the CoordinateCleaner pacakge. Also, the former "species" column is actually the specific epithet of each species, not the species itself, which is "Accepted_name".

data <- data %>% rename(
  "species" = "Accepted_name",
  "epithet" = "Species")

rm(url)
```

# Data cleaning for GDM

First, we need to clean the data.

Remove species with no coordinates and fix coordinate precision problems.

```{r}
data <- data %>%
  filter(!is.na(Latitude) | !is.na(Longitude) ) # remove records with no coordinates

#some problems with coordinate precision:

data$Coordinate.precision <- gsub("0,1 km", "100", data$Coordinate.precision) # 0,1 km is 100 m

data$Coordinate.precision <- gsub("Km", "000", data$Coordinate.precision)

data$Coordinate.precision <- gsub("km", "000", data$Coordinate.precision) # same with capital K

data$Coordinate.precision <- gsub("~100 metres", "100", data$Coordinate.precision) # same with capital K

data$Coordinate.precision <- gsub(" ", "", data$Coordinate.precision) # remove spaces

data$Coordinate.precision <- gsub("<", "", data$Coordinate.precision) # remove < sign. This leaves us with an overestimate, which is good

data$Coordinate.precision <- gsub("&lt;", "", data$Coordinate.precision) #remove "&lt;"

data$Coordinate.precision <- gsub("m", "", data$Coordinate.precision) #remove m

data$Coordinate.precision <- gsub("~", "", data$Coordinate.precision) #remove ~

data$Coordinate.precision <- gsub("gps", NA, data$Coordinate.precision) #remove gps

data$Coordinate.precision <- gsub("GPS; ", NA, data$Coordinate.precision) #remove "GPS; "

data$Coordinate.precision <- gsub("GPS", NA, data$Coordinate.precision) #remove "GPS"

data$Coordinate.precision <- gsub("uncert. < than 100km", NA, data$Coordinate.precision)

data$Coordinate.precision <- gsub("uncert. < than 10km", NA, data$Coordinate.precision)

data$Coordinate.precision <- gsub("uncert. < than 1km", NA, data$Coordinate.precision)

data$Coordinate.precision <- gsub("uncert.than1000", NA, data$Coordinate.precision)

data$Coordinate.precision <- gsub("uncert.than10000", NA, data$Coordinate.precision)


data$Coordinate.precision <- as.numeric(data$Coordinate.precision) # turn into numeric


data <- data %>%
  filter(Coordinate.precision <= 10000 |
           is.na(Coordinate.precision)) # Select coordinates with precision under 10 km or no precision recorded. We will assume that coordinates with no precision have errors lower than 10 km radius.

# this is what we have:
plot(data$Longitude, data$Latitude)
```

Clean coordinates with the *CoordinateCleaner* package.

```{r}
library(CoordinateCleaner)
library(countrycode)

# country names should be in iso3 format, not their English names. Transform:

data$Country <-  countrycode(data$Country, origin =  'country.name', destination = 'iso3c')


# clean coordinates:

geo_flags <- clean_coordinates(x = data,
                               lon = "Longitude",
                               lat = "Latitude",
                               species = "species",
                               countries = "Country",
                               tests = c("capitals",
                                         "centroids",
                                         "equal",
                                         "institutions",
                                         "seas", 
                                         "zeros",
                                         "countries"))

# summary and plot of the problematic cases:
summary(geo_flags)

plot(geo_flags, lon = "Longitude", lat = "Latitude")

# keep only good data:

data <- data[which(geo_flags$.summary == "TRUE"),]
rm(geo_flags)

# Remove countries other than Spain and Portugal:
data <- data %>%
  filter(Country == "ESP" | Country == "PRT") # remove records with no coordinates

# save clean dataset
write.csv2(data, file = "Iberian_bees_clean.csv", row.names = FALSE)

data <- read.csv("Iberian_bees_clean.csv", header = TRUE, sep = ";", dec = ",")

# Remove the balearic islands:
data <- data[-which(data$Province == "Balears, Illes"),]

# also remove records prior than 1980, as our environmental data is 1980-2010

data <- data %>%
  filter(Year <= 1980) 
```

# Environmental data:

Load country shapes:

```{r eval=FALSE, include=FALSE}
es <- getData("GADM", country="ES",level=0, path = "./Data/MAP")

pt <- getData("GADM", country="PT",level=0, path = "./Data/MAP")

iberia <- rbind(es, pt)

#shapefile(iberia, file = "./MAP/Iberia")

rm(es)
rm(pt)
```

Remove Atlantic islands and other places not found in the Iberic mainland with any GIS software.

Load layer again:

```{r, warning=FALSE}
setwd("./Data/MAP")
iberia <- readOGR(dsn = ".", layer = "Iberian_peninsula")
```

Load environmental data and clip it with our map.
Data can be downloaded from https://chelsa-climate.org/downloads/

```{r}
#setwd("../../")

file.list <- list.files("./Data/CHELSA/CURRENT CLIMATE/2.5min/", pattern = "*.tif", full.names = TRUE)

tif.read <- lapply(file.list, raster)

stack.env <- stack(tif.read)

crop.iberia <- crop(stack.env, iberia)
iberia.env.pres <- mask(crop.iberia, iberia)

rm(tif.read)
rm(stack.env)
rm(crop.iberia)
rm(file.list)
```

Normalize environmental layers (to have a mean 0 and standard deviation 1):

```{r}
Normalize(iberia.env.pres)
```

Check colinearity between layers and remove those that are correlated with an R higher than 0.7:

```{r}
iberia.env.pres <- RemCorrLayers(iberia.env.pres, 0.7)
```

Estimate cell sampling stats & filter occurrence data

```{r eval=FALSE, include=FALSE}
# First let's fix abundance data and dates formatting.

data$individualCount <- sum(data$Male, data$Female, data$Worker, data$Not.specified)

data[is.na(data$individualCount),]$individualCount <- 1 # when the number of individuals is not recorded, use a one.

# Create dates for when we have year, month and day complete data. Otherwise, use NA
data$date <- ifelse(!is.na(data$Year) & !is.na(data$Month) & !is.na(data$Day), paste(data$Year, data$Month, data$Day, sep = "-"), NA)

# If there is NA in Year, Month or Day but we have a date in "End.date", keep that, otherwise, keep the original date. 
data$End.date <- gsub("/", "-", data$End.date) # formating
data$End.date <- as.POSIXct(data$End.date, format = "%d-%m-%Y") # formating

data$date <- ifelse(!is.na(data$End.date) & is.na(data$date), data$End.date, data$date)

# Same with "Start.date"
data$Start.date <- gsub("/", "-", data$Start.date)
data$Start.date <- as.POSIXct(data$Start.date, format = "%d-%m-%Y")

data$date <- ifelse(!is.na(data$Start.date) & is.na(data$date), data$Start.date, data$date)

# Keep dates with month and year regardless of they missing day data, but only when there is no start or end date available
data$date <- ifelse(!is.na(data$Year) & !is.na(data$Month) & is.na(data$Day) & is.na(data$date), paste(data$Year, data$Month, data$Day, sep = "-"), data$date)

# Remove data with only year
#data <- data[!is.na(data$date),]

#OOOOR:

# Keep dates with only year regardless of they missing month or day data, but only when there is no start or end date available
#data$date <- ifelse(!is.na(data$Year) & is.na(data$Month) & is.na(data$Day) & is.na(data$date), data$Year, data$date)


# Standardize date formats
data$date <- gsub("/", "-", data$date)
```

# General biodiversity patterns and survey recommendations

```{r}
# Cell stats
data$cell <- cellFromXY(iberia.env.pres, data[, c("Longitude","Latitude")])

cell.stats.iberia <- RichSamp(data, iberia.env.pres, c("Longitude","Latitude","date","species","individualCount","cell"))

#Consider places with at least 10 species recorded as well-sampled

#selected.cells.iberia2 <- cell.stats.iberia$cell[which(cell.stats.iberia$n >= 3 & cell.stats.iberia$Species >= 5)]
selected.cells.iberia <- cell.stats.iberia$cell[which(cell.stats.iberia$Species >= 10)]

#data.sel.iberia2 <- data[which(data$cell %in% selected.cells.iberia2),]
data.sel.iberia <- data[which(data$cell %in% selected.cells.iberia),]
```

Run and map GDM:

```{r}
m1 <- RunGDM(data.sel.iberia, iberia.env.pres, "bray", TRUE, TRUE, c("species", "Longitude", "Latitude"))

summary(m1$gdm.res)

# plot
plotRGB(m1$gdm.map$pcaRast)

#writeRaster(m1$gdm.map$pcaRast, "compositionPCA.TIFF", overwrite = TRUE)

# Plot model fit:

plot(m1$gdm.res, plot.color = "#2274A5", cex = 30, plot.layout=c(3,2))


# Extract I-splines and plot them with axis labels
gdm.1.splineDat <- isplineExtract(m1$gdm.res)

par(mfrow = c(3,2))

plot(gdm.1.splineDat$x[,"bio_1"],
gdm.1.splineDat$y[,"bio_1"],
lwd=3,
type="l",
col = "#2274A5",
xlab="Annual Mean Temperature",
ylab="Partial ecological distance")

plot(gdm.1.splineDat$x[,"bio_3"],
gdm.1.splineDat$y[,"bio_3"],
lwd=3,
type="l",
col = "#2274A5",
xlab="Isothermality",
ylab="Partial ecological distance")

plot(gdm.1.splineDat$x[,"bio_12"],
gdm.1.splineDat$y[,"bio_12"],
lwd=3,
type="l",
col = "#2274A5",
xlab="Annual precipitation",
ylab="Partial ecological distance")

plot(gdm.1.splineDat$x[,"bio_15"],
gdm.1.splineDat$y[,"bio_15"],
lwd=3,
type="l",
col = "#2274A5",
xlab="Precipitation seasonality",
ylab="Partial ecological distance")

# Geographical distance on its own

plot(gdm.1.splineDat$x[,"Geographic"],
gdm.1.splineDat$y[,"Geographic"],
lwd=3,
type="l",
col = "#2274A5",
xlab="Geographic distance",
ylab="Partial ecological distance")
```

# Survey gap analysis:

```{r}
# Generate transformed predictor rasters, based on the raw raster predictor layers
# and the fitted gdm
transRasts <- gdm.transform(model=m1$gdm.res,
data=iberia.env.pres)

# put the values from the transformed layers in a table for easy analysis
Trans.env.table <- as.matrix(transRasts)
col.longs<-xFromCol(transRasts)
row.lats<-yFromRow(transRasts)
Cell_Long<-rep(col.longs, times=nrow(transRasts))
Cell_Lat<-rep(row.lats, each=ncol(transRasts), times=1)
Trans.env.table<-cbind(Cell_Long, Cell_Lat, Trans.env.table)
Trans.env.table <- Trans.env.table[complete.cases(Trans.env.table),]
# specify the locations used to fit the model
ref.coords <- unique(cbind(data.sel.iberia$Longitude, data.sel.iberia$Latitude))
# extract the gdm transformed predictor values for those locations
ref.Trans.env.table <- extract(transRasts,ref.coords)
ref.Trans.env.table <- ref.Trans.env.table[complete.cases(ref.Trans.env.table),]
# Calculate the similarity of each grid cell to the survey cells
# NB - This loop takes a couple of minutes
mean.similarity.region <- rep(0, length=nrow(Trans.env.table))
for(i.cell in 1:nrow(Trans.env.table)){
# Check if this cell has data
if(!is.na(Trans.env.table[i.cell,ncol(Trans.env.table)]))
{
# loop through the reference cells, calculate similarity, add it to the tally
for(j.cell in 1:nrow(ref.Trans.env.table))
{
ecol.dist.1 <- sum(abs(Trans.env.table[i.cell,c(3:ncol(Trans.env.table))] -
ref.Trans.env.table[j.cell,]))
mean.similarity.region[i.cell] <- mean.similarity.region[i.cell] +
(exp(-1 * (m1$gdm.res$intercept + ecol.dist.1)))
} # end for j.cells
# Finish by dividing by the number of neighbouring cells to get the mean
mean.similarity.region[i.cell] <- mean.similarity.region[i.cell] / nrow(ref.Trans.env.table)
} # end if(!is.na())
} # end for i.cell

# Format the similarities into a raster and plot them
mnsim.ras <- raster(transRasts,layer=1)
mnsim.ras <- rasterize(Trans.env.table[,c(1:2)],
                       mnsim.ras,
                       field=mean.similarity.region)

raster:::.plotraster(mnsim.ras,
                    col = heat.colors(100),
                    zlim=c(0,0.3),
                    legend.args = list(text = 'Similarity'))
points(ref.coords)

# Save raster

#writeRaster(mnsim.ras, "SurveyGapAnalysis.TIFF", col = heat.colors(100), overwrite = TRUE)
```

# Survey recommendations:

```{r}
occ.table <- data.sel.iberia

occ.table <- occ.table %>%
  rename(
    "decimalLatitude" = "Latitude",
    "decimalLongitude" = "Longitude"
  )

names(m1$occ.table) <- c("species", "decimalLongitude", "decimalLatitude", "cell")

init <- PreFindNext(m1$gdm.res, m1$occ.table, m1$gdm.rasters, subset.size=1000, search.type="1")

plot(init$out.raster)
points(init$selCoords)

#Find additional survey suggestions

update <- FindNext(init, "add")

plot(update$out.raster)
points(update$selCoords)

update2 <- FindNext(update, "add")

plot(update2$out.raster)
points(update2$selCoords)

update3 <- FindNext(update2, "add")

plot(update3$out.raster)
points(update3$selCoords)

update4 <- FindNext(update3, "add")

plot(update4$out.raster)
points(update4$selCoords)

update5 <- FindNext(update4, "add")

plot(update5$out.raster)
points(update5$selCoords)

update6 <- FindNext(update5, "add")

plot(update6$out.raster)
points(update6$selCoords)

update7 <- FindNext(update6, "add")

plot(update7$out.raster)
points(update7$selCoords)

update8 <- FindNext(update7, "add")

plot(update8$out.raster)
points(update8$selCoords)

update9 <- FindNext(update8, "add")

plot(update9$out.raster)
points(update9$selCoords)

update10 <- FindNext(update9, "add")

plot(update10$out.raster)
points(update10$selCoords)

update11 <- FindNext(update10, "add")

update12 <- FindNext(update11, "add")

update13 <- FindNext(update12, "add")

update14 <- FindNext(update13, "add")

update15 <- FindNext(update14, "add")



# Plot all:
#plotRGB(m1$gdm.map$pcaRast)

raster:::.plotraster(mnsim.ras,
                    col = heat.colors(100),
                    zlim=c(0,0.3),
                    legend.args = list(text = 'Similarity'))
points(update$selCoords, pch = 19, col = "black")
points(update2$selCoords, pch = 19, col = "black")
points(update3$selCoords, pch = 19, col = "black")
points(update4$selCoords, pch = 19, col = "black")
points(update5$selCoords, pch = 19, col = "black")
points(update6$selCoords, pch = 19, col = "black")
points(update7$selCoords, pch = 19, col = "black")
points(update8$selCoords, pch = 19, col = "black")
points(update9$selCoords, pch = 19, col = "black")
points(update10$selCoords, pch = 19, col = "black")
points(update11$selCoords, pch = 19, col = "black")
points(update12$selCoords, pch = 19, col = "black")
points(update13$selCoords, pch = 19, col = "black")
points(update14$selCoords, pch = 19, col = "black")
points(update15$selCoords, pch = 19, col = "black")
```

# Future projection

How big will be the changes provoked by climate change to these points in which sampling effort should increase?

First load all future data.

```{r}
# 2010-2040 ssp126
file.list <- list.files("./Data/CHELSA/FUTURE CLIMATE/2010-2040/ssp126/2.5min", pattern = "*.tif", full.names = TRUE)

tif.read <- lapply(file.list, raster)

future.20102040ssp126 <- stack(tif.read)

crop.iberia.fut <- crop(future.20102040ssp126, iberia)
iberia.fut.20102040ssp126 <- mask(crop.iberia.fut, iberia)

Normalize(iberia.fut.20102040ssp126)


# 2010-2040 ssp370
file.list <- list.files("./Data/CHELSA/FUTURE CLIMATE/2010-2040/ssp370/2.5min", pattern = "*.tif", full.names = TRUE)

tif.read <- lapply(file.list, raster)

future.20102040ssp370 <- stack(tif.read)

crop.iberia.fut <- crop(future.20102040ssp370, iberia)
iberia.fut.20102040ssp370 <- mask(crop.iberia.fut, iberia)

Normalize(iberia.fut.20102040ssp370)


# 2010-2040 ssp585
file.list <- list.files("./Data/CHELSA/FUTURE CLIMATE/2010-2040/ssp585/2.5min", pattern = "*.tif", full.names = TRUE)

tif.read <- lapply(file.list, raster)

future.20102040ssp585 <- stack(tif.read)

crop.iberia.fut <- crop(future.20102040ssp585, iberia)
iberia.fut.20102040ssp585 <- mask(crop.iberia.fut, iberia)

Normalize(iberia.fut.20102040ssp585)


# 2041-2070 ssp126
file.list <- list.files("./Data/CHELSA/FUTURE CLIMATE/2041-2070/ssp126/2.5min", pattern = "*.tif", full.names = TRUE)

tif.read <- lapply(file.list, raster)

future.20412070ssp126 <- stack(tif.read)

crop.iberia.fut <- crop(future.20412070ssp126, iberia)
iberia.fut.20412070ssp126 <- mask(crop.iberia.fut, iberia)

Normalize(iberia.fut.20412070ssp126)


# 2041-2070 ssp370
file.list <- list.files("./Data/CHELSA/FUTURE CLIMATE/2041-2070/ssp370/2.5min", pattern = "*.tif", full.names = TRUE)

tif.read <- lapply(file.list, raster)

future.20412070ssp370 <- stack(tif.read)

crop.iberia.fut <- crop(future.20412070ssp370, iberia)
iberia.fut.20412070ssp370 <- mask(crop.iberia.fut, iberia)

Normalize(iberia.fut.20412070ssp370)


# 2041-2070 ssp585
file.list <- list.files("./Data/CHELSA/FUTURE CLIMATE/2041-2070/ssp585/2.5min", pattern = "*.tif", full.names = TRUE)

tif.read <- lapply(file.list, raster)

future.20412070ssp585 <- stack(tif.read)

crop.iberia.fut <- crop(future.20412070ssp585, iberia)
iberia.fut.20412070ssp585 <- mask(crop.iberia.fut, iberia)

Normalize(iberia.fut.20412070ssp585)


# 2071-2100 ssp126
file.list <- list.files("./Data/CHELSA/FUTURE CLIMATE/2071-2100/ssp126/2.5min", pattern = "*.tif", full.names = TRUE)

tif.read <- lapply(file.list, raster)

future.20712100ssp126 <- stack(tif.read)

crop.iberia.fut <- crop(future.20712100ssp126, iberia)
iberia.fut.20712100ssp126 <- mask(crop.iberia.fut, iberia)

Normalize(iberia.fut.20712100ssp126)


# 2071-2100 ssp370
file.list <- list.files("./Data/CHELSA/FUTURE CLIMATE/2071-2100/ssp370/2.5min", pattern = "*.tif", full.names = TRUE)

tif.read <- lapply(file.list, raster)

future.20712100ssp370 <- stack(tif.read)

crop.iberia.fut <- crop(future.20712100ssp370, iberia)
iberia.fut.20712100ssp370 <- mask(crop.iberia.fut, iberia)

Normalize(iberia.fut.20712100ssp370)


# 2071-2100 ssp585
file.list <- list.files("./Data/CHELSA/FUTURE CLIMATE/2071-2100/ssp585/2.5min", pattern = "*.tif", full.names = TRUE)

tif.read <- lapply(file.list, raster)

future.20712100ssp585 <- stack(tif.read)

crop.iberia.fut <- crop(future.20712100ssp585, iberia)
iberia.fut.20712100ssp585 <- mask(crop.iberia.fut, iberia)

Normalize(iberia.fut.20712100ssp585)

rm(tif.read)
rm(file.list)  
```
Project to future scenarios:

```{r}
# 2040

iberia.env.pres.sel <- iberia.env.pres[[which(names(iberia.env.pres) %in% m1$gdm.res$predictors)]]

iberia.fut.2040.126.sel <- iberia.fut.20102040ssp126[[which(names(iberia.fut.20102040ssp126) %in% m1$gdm.res$predictors)]]

timePred.2040.126 <- predict(object=m1$gdm.res,
data=iberia.env.pres.sel,
time=T,
predRasts=iberia.fut.2040.126.sel)

plot(timePred.2040.126)

iberia.fut.2040.370.sel <- iberia.fut.20102040ssp370[[which(names(iberia.fut.20102040ssp370) %in% m1$gdm.res$predictors)]]

timePred.2040.370 <- predict(object=m1$gdm.res,
data=iberia.env.pres.sel,
time=T,
predRasts=iberia.fut.2040.370.sel)

plot(timePred.2040.370)

iberia.fut.2040.585.sel <- iberia.fut.20102040ssp585[[which(names(iberia.fut.20102040ssp585) %in% m1$gdm.res$predictors)]]

timePred.2040.585 <- predict(object=m1$gdm.res,
data=iberia.env.pres.sel,
time=T,
predRasts=iberia.fut.2040.585.sel)

plot(timePred.2040.585)

# 2070

iberia.fut.2070.126.sel <- iberia.fut.20412070ssp126[[which(names(iberia.fut.20412070ssp126) %in% m1$gdm.res$predictors)]]

timePred2070.126 <- predict(object=m1$gdm.res,
data=iberia.env.pres.sel,
time=T,
predRasts=iberia.fut.2070.126.sel)

plot(timePred2070.126)

iberia.fut.2070.370.sel <- iberia.fut.20412070ssp370[[which(names(iberia.fut.20412070ssp370) %in% m1$gdm.res$predictors)]]

timePred2070.370 <- predict(object=m1$gdm.res,
data=iberia.env.pres.sel,
time=T,
predRasts=iberia.fut.2070.370.sel)

plot(timePred2070.370)

iberia.fut.2070.585.sel <- iberia.fut.20412070ssp585[[which(names(iberia.fut.20412070ssp585) %in% m1$gdm.res$predictors)]]

timePred2070.585 <- predict(object=m1$gdm.res,
data=iberia.env.pres.sel,
time=T,
predRasts=iberia.fut.2070.585.sel)

plot(timePred2070.585)

# 2100

iberia.fut.2100.126.sel <- iberia.fut.20712100ssp126[[which(names(iberia.fut.20712100ssp126) %in% m1$gdm.res$predictors)]]

timePred2100.126 <- predict(object=m1$gdm.res,
data=iberia.env.pres.sel,
time=T,
predRasts=iberia.fut.2100.126.sel)

plot(timePred2100.126)

iberia.fut.2100.370.sel <- iberia.fut.20712100ssp370[[which(names(iberia.fut.20712100ssp370) %in% m1$gdm.res$predictors)]]

timePred2100.370 <- predict(object=m1$gdm.res,
data=iberia.env.pres.sel,
time=T,
predRasts=iberia.fut.2100.370.sel)

plot(timePred2100.370)

iberia.fut.2100.585.sel <- iberia.fut.20712100ssp585[[which(names(iberia.fut.20712100ssp585) %in% m1$gdm.res$predictors)]]

timePred2100.585 <- predict(object=m1$gdm.res,
data=iberia.env.pres.sel,
time=T,
predRasts=iberia.fut.2100.585.sel)

plot(timePred2100.585)







raster:::.plotraster(timePred2100.126,
                    col = rev(heat.colors(100)),
                    zlim=c(0.7,1),
                    legend.args = list(text = 'Turnover'))
```

Extract values from the survey recommendations:

```{r}
survey.recomm <- rbind(update$selCoords, update2$selCoords, update3$selCoords, update4$selCoords, update5$selCoords, update6$selCoords, update7$selCoords, update8$selCoords, update9$selCoords, update10$selCoords, update11$selCoords, update12$selCoords, update13$selCoords, update14$selCoords, update15$selCoords)

survey2040126 <- raster::extract(timePred.2040.126, survey.recomm)

survey2040370 <- raster::extract(timePred.2040.370, survey.recomm)

survey2040585 <- raster::extract(timePred.2040.585, survey.recomm)

survey2070126 <- raster::extract(timePred2070.126, survey.recomm)

survey2070370 <- raster::extract(timePred2070.370, survey.recomm)

survey2070585 <- raster::extract(timePred2070.585, survey.recomm)

survey2100126 <- raster::extract(timePred2100.126, survey.recomm)

survey2100370 <- raster::extract(timePred2100.370, survey.recomm)

survey2100585 <- raster::extract(timePred2100.585, survey.recomm)


# compare with our currently sampled grids:

data.sel.iberia.coords <- data.frame(data.sel.iberia$Latitude, data.sel.iberia$Longitude, data.sel.iberia$cell)

# Keep one coordinate per cell:

data.sel.iberia.coords <- data.sel.iberia.coords %>% 
  group_by(data.sel.iberia.cell) %>%
  summarize(avg_latitude = mean(data.sel.iberia.Latitude),
            avg_longitude = mean(data.sel.iberia.Longitude))

names(data.sel.iberia.coords) <- c("cell", "y", "x")

# extract:

surveyed2040126 <- raster::extract(timePred.2040.126, data.sel.iberia.coords[,3:2])

surveyed2040370 <- raster::extract(timePred.2040.370, data.sel.iberia.coords[,3:2])

surveyed2040585 <- raster::extract(timePred.2040.585, data.sel.iberia.coords[,3:2])

surveyed2070126 <- raster::extract(timePred2070.126, data.sel.iberia.coords[,3:2])

surveyed2070370 <- raster::extract(timePred2070.370, data.sel.iberia.coords[,3:2])

surveyed2070585 <- raster::extract(timePred2070.585, data.sel.iberia.coords[,3:2])

surveyed2100126 <- raster::extract(timePred2100.126, data.sel.iberia.coords[,3:2])

surveyed2100370 <- raster::extract(timePred2100.370, data.sel.iberia.coords[,3:2])

surveyed2100585 <- raster::extract(timePred2100.585, data.sel.iberia.coords[,3:2])

# merge sampled and unsampled cells
survey2040126 <- data.frame(value = survey2040126, group = rep("recommendation",15))
surveyed2040126 <- data.frame(value = surveyed2040126, group = rep("surveyed",15))

fut.diff2040126 <- rbind(survey2040126, surveyed2040126)

fut.diff2040126$year <- 2040
fut.diff2040126$SSP <- 126
fut.diff2040126$coord <- rbind(survey.recomm, data.sel.iberia.coords[,3:2])

survey2040370 <- data.frame(value = survey2040370, group = rep("recommendation",15))
surveyed2040370 <- data.frame(value = surveyed2040370, group = rep("surveyed",15))

fut.diff2040370 <- rbind(survey2040370, surveyed2040370)

fut.diff2040370$year <- 2040
fut.diff2040370$SSP <- 370
fut.diff2040370$coord <- rbind(survey.recomm, data.sel.iberia.coords[,3:2])

survey2040585 <- data.frame(value = survey2040585, group = rep("recommendation",15))
surveyed2040585 <- data.frame(value = surveyed2040585, group = rep("surveyed",15))

fut.diff2040585 <- rbind(survey2040585, surveyed2040585)

fut.diff2040585$year <- 2040
fut.diff2040585$SSP <- 585
fut.diff2040585$coord <- rbind(survey.recomm, data.sel.iberia.coords[,3:2])

survey2070126 <- data.frame(value = survey2070126, group = rep("recommendation",15))
surveyed2070126 <- data.frame(value = surveyed2070126, group = rep("surveyed",15))

fut.diff2070126 <- rbind(survey2070126, surveyed2070126)

fut.diff2070126$year <- 2070
fut.diff2070126$SSP <- 126
fut.diff2070126$coord <- rbind(survey.recomm, data.sel.iberia.coords[,3:2])

survey2070370 <- data.frame(value = survey2070370, group = rep("recommendation",15))
surveyed2070370 <- data.frame(value = surveyed2070370, group = rep("surveyed",15))

fut.diff2070370 <- rbind(survey2070370, surveyed2070370)

fut.diff2070370$year <- 2070
fut.diff2070370$SSP <- 370
fut.diff2070370$coord <- rbind(survey.recomm, data.sel.iberia.coords[,3:2])

survey2070585 <- data.frame(value = survey2070585, group = rep("recommendation",15))
surveyed2070585 <- data.frame(value = surveyed2070585, group = rep("surveyed",15))

fut.diff2070585 <- rbind(survey2070585, surveyed2070585)

fut.diff2070585$year <- 2070
fut.diff2070585$SSP <- 585
fut.diff2070585$coord <- rbind(survey.recomm, data.sel.iberia.coords[,3:2])

survey2100126 <- data.frame(value = survey2100126, group = rep("recommendation",15))
surveyed2100126 <- data.frame(value = surveyed2100126, group = rep("surveyed",15))

fut.diff2100126 <- rbind(survey2100126, surveyed2100126)

fut.diff2100126$year <- 2100
fut.diff2100126$SSP <- 126
fut.diff2100126$coord <- rbind(survey.recomm, data.sel.iberia.coords[,3:2])

survey2100370 <- data.frame(value = survey2100370, group = rep("recommendation",15))
surveyed2100370 <- data.frame(value = surveyed2100370, group = rep("surveyed",15))

fut.diff2100370 <- rbind(survey2100370, surveyed2100370)

fut.diff2100370$year <- 2100
fut.diff2100370$SSP <- 370
fut.diff2100370$coord <- rbind(survey.recomm, data.sel.iberia.coords[,3:2])

survey2100585 <- data.frame(value = survey2100585, group = rep("recommendation",15))
surveyed2100585 <- data.frame(value = surveyed2100585, group = rep("surveyed",15))

fut.diff2100585 <- rbind(survey2100585, surveyed2100585)

fut.diff2100585$year <- 2100
fut.diff2100585$SSP <- 585
fut.diff2100585$coord <- rbind(survey.recomm, data.sel.iberia.coords[,3:2])

# merge all future scenarios

all.data <- rbind(fut.diff2040126, fut.diff2040370, fut.diff2040585, fut.diff2070126, fut.diff2070370, fut.diff2070585, fut.diff2100126, fut.diff2100370, fut.diff2100585)

all.data$year <- as.factor(all.data$year) #factorize
all.data$SSP <- as.factor(all.data$SSP) #factorize

all.data$site <- paste0(all.data$coord$x, all.data$coord$y, sep = "") # create a variable to use as random factor

all.data$yearssp <- paste0(all.data$year, all.data$SSP, sep = "") # create a variable for each scenario (fut+ssp)


# Run GLMM:

# Transform biological change variable with arcsin(sqrt())

all.data2 <- all.data # backup copy
all.data2$value <- asin(sqrt(all.data2$value)) #transformation

glmm <- glmmTMB(value ~ group*yearssp + (1|site), data = all.data2, family = Gamma(link=identity), na.action = "na.exclude")

summary(glmm)

AIC(glmm)
sim_output <- simulateResiduals(fittedModel = glmm, plot = F) # simulate residuals
plot(sim_output) # model evaluation

glmmTMB:::Anova.glmmTMB(
  glmm,
  type = ("III"),
  test.statistic = ("Chisq"),
  component = "cond",
  vconv. = vcov(glmm)[["cond"]]
)

# Save coordinates:
write.csv(survey.recomm, "survey_recommendations.csv", quote = FALSE, row.names = FALSE)

write.csv(ref.coords, "wellsampled_cells.csv", quote = FALSE, row.names = FALSE)


library(ggplot2)

ggplot(all.data) +
 aes(x = value, y = SSP, fill = group) +
 geom_boxplot() +
 scale_fill_hue(direction = 1) +
 theme_minimal() +
 facet_wrap(vars(year))


all.data2[all.data2$group == "recommendation",]$group <- "unsurveyed"

ggplot(all.data2) +
  aes(x = value, y = SSP, fill = group) +
  geom_boxplot() +
  scale_fill_manual(
    values = c(unsurveyed = "#FF8600",
    surveyed = "#005900")
  ) +
  labs(x = "Expected biological change") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  facet_wrap(vars(year)) +
  coord_flip()
```
